\documentclass[11pt]{article}
\usepackage[margin=1.1in]{geometry}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage[hidelinks]{hyperref}
\usepackage{listings}
\usepackage{xcolor}
\lstset{basicstyle=\ttfamily\small, breaklines=true, columns=fullflexible}

% Every number below comes from generated/ — built by `go run ./cmd/bench
% paper <run dirs>` from the committed episode evidence. Do not hand-edit
% figures or tables.
\input{generated/summary_macros}

\title{Compiler-Validated Semantic Editing for Weak Local Coding Models}
\author{Guy J.\ Grigsby\\\small\texttt{guy@grigsby.dev}}
\date{Draft — \today}

\begin{document}
\maketitle

\begin{abstract}
Coding agents edit source code as text, and small locally runnable
models fail at it: diff formats break, identifiers get guessed, and
errors surface only after files change. ago is a semantic edit protocol
for Go. The agent queries a typechecked workspace and submits mutations
from a versioned operation catalog; the compiler validates every
mutation against an in-memory snapshot before anything reaches disk.
Edits apply as atomic, generation-checked transactions or not at all,
so an edit that does not typecheck is unrepresentable, not merely
checked. Rejections are the conversation: each carries typed
diagnostics and paste-ready repairs (the corrected call, verbatim), and
resending a rejected call unchanged escalates. An oracle replays each
task's ground-truth commit through the protocol before any model
attempts it, so a task enters the benchmark only once it is proven
solvable; tasks are mined from real commits in traefik, vault,
boundary, and cobra. Across \EpisodeCount{} recorded episodes, the
semantic mode passes \SemanticPassRate{} versus \RawPassRate{} for raw
file editing under identical prompts and budgets, and the gap widens as
models shrink: a quantized 9B model passes 75\% of episodes through the
protocol against 45\% raw, with raw at zero on five of eleven tasks the
protocol completes. All code, per-episode evidence (transcripts,
request logs, serving configurations), and analysis tooling are
released.
\end{abstract}

\section{Introduction}
% Thesis: weak local models become effective repo-scale Go editors when
% raw file editing is replaced by semantic queries and validated
% mutations with typed rejections. Expand from idea.md; keep the
% conversational-repair framing central: the contribution is not an op
% catalog, it is an interaction design under which weak models recover.
TODO(guy): expand from idea.md.

\section{Related Work}
% From the 2026-07-16 landscape survey; expand with proper citations.
Symbol-level agent tooling over LSP (Serena; gopls's experimental MCP
server) navigates semantically but still edits text, or exposes a
single mutation. The compiler diagnostics sit right there in those
stacks; the difference here is that validation is not advisory.
CODESTRUCT frames agents over structured action spaces with
syntax-validated edits and reports token and accuracy gains; ago's
mutations are type-checked, transactional, and carry repairs.
Editing-format benchmarks (EDIT-Bench, RES-Q, FEA-Bench) measure text
formats and find that strict formats hurt current models. That is the
failure mode the repair channel converts: the format stays strict and
the rejection hands back the corrected call. Compiler-guided patch
generation (Repilot) and typed-hole context (Hazel) take the same
validate-during-generation stance at expression granularity.
TODO: full citations.

\section{The ago Protocol}
\subsection{Surface}
Ten tools carry a versioned catalog of 33 operations across
declaration, statement, test, and project families, plus seven query
kinds with deterministic, paged responses. Node handles address
statements inside declarations; generation counters make staleness
explicit instead of silent.
\subsection{Transactions}
An ordered op list validates as one unit against an in-memory snapshot.
Ops apply to a copy, the dirty set re-typechecks once (in parallel,
scheduled over post-edit import edges), resolution proofs run (rename
rejects reference capture the compiler would happily accept), then all
files write together or nothing does. Pre-existing issues in the
affected packages are baselined and tolerated; real codebases carry
history, and only new diagnostics reject.
\subsection{Rejection as conversation}
Every rejection answers one question: what is the correct next call.
It carries typed diagnostics, candidate lists, and
\texttt{possible\_repairs}, complete calls the agent resends verbatim.
Repairs are tested by execution before they are offered; a repair that
would itself reject is worse than none. Resending a rejected call
unchanged escalates. Accepted mutations embed the fresh view of the
touched declaration, so back-to-back edits skip the read round trip.

\section{Oracle Methodology}
Every benchmark task comes from a real commit, so the correct answer is
already known. Before any model attempts a task, the oracle submits
that known answer through the protocol itself. If the right answer
cannot get through, the model never had a chance: the problem is ours
(an engine bug, an extractor error, or a named protocol ceiling), and
the task stays out of the roster until it is fixed. This separates
``the model failed'' from ``the task was impossible'' by construction,
and the oracle transcripts double as a training corpus.

\section{Benchmark}
Same model, same harness, same prompts (token parity is enforced by
test), same time cap. One arm edits files with shell tools; the other
sees only the protocol. Scoring is a goal predicate plus a clean
typecheck plus scoped tests, and the test gate is voided where a
pristine worktree also fails it. Serving configurations are pinned per
run. Every episode records its transcript, diff, score, token usage,
failure kind, and the daemon's request log; the evidence is the
contribution as much as the numbers.

\section{Results}
\begin{table}[h]
\centering
\resizebox{\textwidth}{!}{\input{generated/results_table}}
\caption{Per-task, per-mode outcomes. Generated from committed episode
evidence; pass@k over k episodes, median time-to-green over passing
episodes.}
\end{table}
Semantic passed \SemanticPassCount{} episodes; raw passed
\RawPassCount{}. The per-task table tells the sharper story: raw wins
only where grep suffices, and goes to zero on the tasks that need
whole-workspace consistency.
TODO: repair-uptake analysis (the 9B model recovers through rejections
it is offered no repairs for; the 30B-class model ignores repairs it
is offered), ablation arm, floor-model section, the cfff8d42 case
study (discovery-primitive gap found and fixed).

\section{Limitations}
One language. Contamination is held constant across arms but inflates
absolute rates. One serving stack per model, small k, and the
infrastructure is author-run. TODO: expand honestly.

\section{Availability}
Code (MIT), all episode evidence, and analysis tooling:
\url{https://github.com/guygrigsby/agent-go}.

\end{document}
